---
title: "Exploratory analysis script by Marco Reina"
output: html_document
---

```{r}
# Loading libraries
library(readr)
library(here)
library(dplyr)
library(countries)
library(ggplot2)
```

### Raw data seems too big to ensure reproducibility. Smaller dataframes were created from the raw data. EDA will start from those files. The code to create the initial files is the following:

```r
#Isolates obtained from _NCBI Pathogen Detection_ database: Salmonella
isolates_raw <- read_tsv(here("data/raw-data", "isolates.tsv"))

#Changing format from tsv to csv
write_csv(isolates_raw, here("data/processed-data/", "isolates1.csv"))

#Making a new object based on new csv
isolates0 <- read.csv(here("data/processed-data/", "isolates1.csv"))

#Genomes dataset from NCBI: _Bacterial genomes_
genomes_raw <- read_tsv(here("data/raw-data/", "genomes_datasets.tsv"))

#Changing format from tsv to csv
write_csv(genomes_raw, here("data/processed-data/", "genomes1.csv"))

#Making a new object based on new csv
genomes0 <- read.csv(here("data/processed-data/", "genomes1.csv"))

#Removing non useful columns from isolates0
isolates0$WGS.accession	=	NULL
isolates0$WGS.prefix	=	NULL
isolates0$PFGE.secondary.enzyme.pattern	=	NULL
isolates0$PFGE.primary.enzyme.pattern	=	NULL
isolates0$Platform	=	NULL
isolates0$Host.disease	=	NULL
isolates0$Outbreak	=	NULL
isolates0$SRA.release.date	=	NULL
isolates0$SRA.Center	=	NULL
isolates0$Run	=	NULL
isolates0$Library.layout	=	NULL
isolates0$Lat.Lon	=	NULL
isolates0$AST.phenotypes	=	NULL
isolates0$AMR.genotypes.core	=	NULL
isolates0$Host	=	NULL
isolates0$Min.same	=	NULL
isolates0$Min.diff	=	NULL

#Removing non-useful columns from genomes0
genomes0$Assembly.Paired.Assembly.Accession	=	NULL
genomes0$Organism.Infraspecific.Names.Breed	=	NULL
genomes0$Organism.Infraspecific.Names.Strain	=	NULL
genomes0$Organism.Infraspecific.Names.Cultivar	=	NULL
genomes0$Organism.Infraspecific.Names.Ecotype	=	NULL
genomes0$Organism.Infraspecific.Names.Isolate	=	NULL
genomes0$Organism.Infraspecific.Names.Sex	=	NULL
genomes0$WGS.project.accession	=	NULL
genomes0$Annotation.BUSCO.Complete	=	NULL
genomes0$Annotation.BUSCO.Single.Copy	=	NULL
genomes0$Annotation.BUSCO.Duplicated	=	NULL
genomes0$Annotation.BUSCO.Fragmented	=	NULL
genomes0$Annotation.BUSCO.Missing	=	NULL
genomes0$Annotation.BUSCO.Lineage	=	NULL
genomes0$Type.Material.Display.Text	=	NULL
genomes0$CheckM.marker.set	=	NULL
genomes0$CheckM.completeness	=	NULL
genomes0$CheckM.contamination	=	NULL
genomes0$slm_filter	=	NULL
genomes0$release_year	=	NULL

#Creating BioSample column
genomes0$BioSample	=	genomes0$Assembly.BioSample.Accession

#Both dataframes are still too large >100 MB they will be streamlined even more by keeping only selected columns

#Selected columns for isolates dataframe
isolates1 <- data.frame("Create.date"	=	isolates0$Create.date,
                        "Location"	=	isolates0$Location,
                        "Source.type"	=	isolates0$Source.type,
                        "BioSample"	=	isolates0$BioSample,
                        "AMR.genotypes"	=	isolates0$AMR.genotypes,
                        "Computed.types"	=	isolates0$Computed.types
)

isolates1 <- na.omit(isolates1)

#Identifying country names where isolates came from to NCBI
isolates1$country_name <- country_name(x= isolates1$Location, to="ISO3")

#Subsetting only for isolates obtained in the USA in the last 5 complete years
isolates1 <- isolates1[(isolates1$country_name %in% "USA"), ]

isolates1$year <-  as.numeric(substr(isolates1$Create.date, 1, 4))

isolates1 <- isolates1[(isolates1$year %in% c(2025, 2024, 2023, 2022, 2021)), ]

#Obtaining only categories of interest
isolates1 <- isolates1[(isolates1$Source.type %in% c("animal", "enviromental", "food", "human")), ]

#Saving new df, file under 100 MB
write_csv(isolates1, here("data/processed-data/", "isolates1.csv"))

#Selected colums for genomes dataframe
genomes0$genus <-  substr(genomes0$Organism.Name, 1, 10)

genomes1 <- data.frame("Annotation.Release.Date" = genomes0$Annotation.Release.Date,
                        "Assembly.Stats.Total.Sequence.Length" = genomes0$Assembly.Stats.Total.Sequence.Length,
                        "Assembly.Stats.Number.of.Contigs" = genomes0$Assembly.Stats.Number.of.Contigs,
                        "Assembly.Level" = genomes0$Assembly.Level,
                        "Assembly.Release.Date" = genomes0$Assembly.Release.Date,
                        "BioSample" = genomes0$BioSample,
                        "Annotation.Count.Gene.Total" = genomes0$Annotation.Count.Gene.Total,
                        "Annotation.Count.Gene.Protein.coding" = genomes0$Annotation.Count.Gene.Protein.coding,
                        "Annotation.Count.Gene.Pseudogene" = genomes0$Annotation.Count.Gene.Pseudogene,
                        "genus" = genomes0$genus 
)

genomes1 <- genomes1[(genomes1$genus %in% "Salmonella"), ]

genomes1 <- na.omit(genomes1)

write_csv(genomes1, here("data/processed-data/", "genomes1.csv"))
```

### After creating files for use 
# EDA CODE STARTS HERE:
```{r}
#Load files created
isolates1 <- read.csv(here("data/processed-data/", "isolates1.csv"))
summary(isolates1)

genomes1 <- read.csv(here("data/processed-data/", "genomes1.csv"))
summary(genomes1)
```

### Exploring isolates1.csv
#### Data comes from progam of _Pathogen Detection_ from NCBI
```{r}
#Observing if there is any trend in the years the isolates were obtained
ggplot(isolates1, aes(x = year)) + geom_bar()

#Observing trends main within isolates
ggplot(isolates1, aes(x = year, fill = Source.type)) + geom_bar()

ggplot(isolates1, aes(x = year)) + geom_bar() + facet_wrap(~ Source.type, scales = "free", ncol = 1, strip.position = "right")
```

### Exploring genomes1.csv
#### Data comes from _Datasets_ by NIH
```{r}
#Substracting year of release date
genomes1$year <- as.numeric(substr(genomes1$Assembly.Release.Date , 1, 4))

#Observing trends main within genomes
ggplot(genomes1, aes(x = year)) + geom_bar()

#Trends of available genomes from the last 5 complete years
genomes2 <- genomes1[(genomes1$year %in% c(2025, 2024, 2023, 2022, 2021)), ]
ggplot(genomes2, aes(x = year)) + geom_bar()

#Histograms for all rest of the parameter
ggplot(genomes2, aes(x = Assembly.Stats.Total.Sequence.Length)) + geom_histogram()

ggplot(genomes2, aes(x = Assembly.Stats.Number.of.Contigs)) + geom_histogram()

ggplot(genomes2, aes(x = Annotation.Count.Gene.Total)) + geom_histogram()

ggplot(genomes2, aes(x = Annotation.Count.Gene.Protein.coding)) + geom_histogram()

ggplot(genomes2, aes(x = Annotation.Count.Gene.Pseudogene)) + geom_histogram()

ggplot(genomes2, aes(x = Assembly.Level)) + geom_bar()
```

### Merging both dataframes to assign sources, AMR genotype, and computed serovar
```{r}
#First merge based on BioSample column only samples that intersect in both databases
merged_df0 <- inner_join(genomes2, isolates1, relationship = "many-to-many",  keep = FALSE, by = "BioSample")

#Only 197 rows match between genomes and isolates
merged_df0 <- merged_df0 %>% 
  mutate(Annotation.Release.Date = as.Date(Annotation.Release.Date))

#Removing repeated observations as they are several annotations dates, filtered by newer
merged_df1 <- merged_df0 %>% 
  group_by(BioSample) %>% 
  filter(Annotation.Release.Date== max(Annotation.Release.Date, na.rm = TRUE)) %>% 
  ungroup()

#Saving file
write_csv(merged_df1, here("data/processed-data/", "merged_df1.csv"))
```

### Exploring merged_df1.csv
```{r}
summary(merged_df1)

ggplot(merged_df1, aes(x = year.y, fill = Source.type)) + geom_bar()

ggplot(merged_df1, aes(x = year.y)) + geom_bar() + facet_wrap(~ Source.type, scales = "free", ncol = 1, strip.position = "right")

ggplot(merged_df1, aes(x = Assembly.Stats.Total.Sequence.Length)) + geom_histogram()

ggplot(merged_df1, aes(x = Assembly.Stats.Number.of.Contigs)) + geom_histogram()

ggplot(merged_df1, aes(x = Annotation.Count.Gene.Total)) + geom_histogram()

ggplot(merged_df1, aes(x = Annotation.Count.Gene.Protein.coding)) + geom_histogram()

ggplot(merged_df1, aes(x = Annotation.Count.Gene.Pseudogene)) + geom_histogram()

ggplot(merged_df1, aes(x = Assembly.Level)) + geom_bar()

#Filtering by only complete genomes
merged_df2 <- merged_df1[(merged_df1$Assembly.Level %in% "Complete Genome"), ]

summary(merged_df2)

#Repeating exploration
ggplot(merged_df2, aes(x = year.y, fill = Source.type)) + geom_bar()

ggplot(merged_df2, aes(x = year.y)) + geom_bar() + facet_wrap(~ Source.type, scales = "free", ncol = 1, strip.position = "right")

ggplot(merged_df2, aes(x = Assembly.Stats.Total.Sequence.Length)) + geom_histogram()

ggplot(merged_df2, aes(x = Assembly.Stats.Number.of.Contigs)) + geom_histogram()

ggplot(merged_df2, aes(x = Annotation.Count.Gene.Total)) + geom_histogram()

ggplot(merged_df2, aes(x = Annotation.Count.Gene.Protein.coding)) + geom_histogram()

ggplot(merged_df2, aes(x = Annotation.Count.Gene.Pseudogene)) + geom_histogram()
```

### Only 137 genomes are complete and intersect between both datasets. These will be use only to assess genome structure
### Isolates1 will be use to assess AMR genotypes
### Genomes1-3 dataframes resulted to be only complimentary for the analysis 
