---
title: "Descriptive analysis script by Marco Reina"
output: html_document
---
#### Code originally written by MR and improve with ChatGPT & Qween

```{r}
# Load libraries
library(tidyverse)
library(readr)
library(here)
library(dplyr)
library(stringr)
library(knitr)
library(forcats) 
library(tidytext)

# Read data
isolates1 <- read_csv(here("data/processed-data/", "isolates1.csv"))

# Get the total number of isolates
totaln_isolates <- nrow(isolates1)

#Split computed serovars
isolates1 <- isolates1 %>%
  mutate(
    antigen_formula = str_extract(Computed.types, "(?<=antigen_formula=).*?(?=,serotype=)"),
    serotype = str_extract(Computed.types, "(?<=serotype=).*$")
  )

# Removing columns that are no longer useful after the original cleaning
isolates1$Computed.types	=	NULL
isolates1$Location	=	NULL
isolates1$country_name	=	NULL 
```

##### The goal of this analysis is to characterize AMR incidence in _Salmonella_
##### As the data comes from an observational surveillance program, only descriptive analysis will be performed. This approach has changed from the original intention described in the manuscript.

```{r}
#Split AMR genes
amr_long <- isolates1 %>%
  # Keep isolates even if no AMR genes
  mutate(AMR.genotypes = if_else(
    is.na(AMR.genotypes) | AMR.genotypes == "",
    "None=NONE",
    AMR.genotypes
  )) %>%
  # Split genes into rows
  separate_rows(AMR.genotypes, sep = ",") 
```

```{r}
# Assign antimicrobial class

amr_long <- amr_long %>%
   mutate(
      AMR_class = case_when(
      # === 1. No AMR ===
      str_detect(AMR.genotypes, "None") ~ "No AMR",
      
      # === 2. Beta-lactams ===
      str_detect(AMR.genotypes, "^bla") ~ "Beta_lactam",
      
      # === 3. Tetracyclines ===
      str_detect(AMR.genotypes, "tet") ~ "Tetracycline",
      
      # === 4. Sulfonamides ===
      str_detect(AMR.genotypes, "^sul") ~ "Sulfonamide",
      
      # === 5. Trimethoprim ===
      str_detect(AMR.genotypes, "^dfr") ~ "Trimethoprim",
      
      # === 6. Quinolones: target mutations & protection proteins ===
      str_detect(AMR.genotypes, "gyrA|gyrB|parC_|parE_|^qnr|^oqx|^qepA") ~ "Quinolone",
      
      # === 7. Aminoglycosides: modifying enzymes (broadened) ===
      str_detect(AMR.genotypes, "aad|aph\\(|aac\\(|armA|^rmt|^ant|^aphA|^aacA") ~ "Aminoglycoside",
      
      # === 8. Phenicols (chloramphenicol family) ===
      str_detect(AMR.genotypes, "floR|^cat|^cml|^gar") ~ "Phenicol",  
      
      # === 9. Macrolides: including esterases (estT/estX) ===
      str_detect(AMR.genotypes, "erm|^mef|^mph|^msr|^ere|^est") ~ "Macrolide",  

      # === 10. Lincosamides ===
      str_detect(AMR.genotypes, "^lnu") ~ "Lincosamide",
      
      # === 11. Fosfomycin ===
      str_detect(AMR.genotypes, "^fos") ~ "Fosfomycin",
      
      # === 12. Colistin/Polymyxin: mcr enzymes + pmr regulators ===
      str_detect(AMR.genotypes, "^mcr|^pmrA_|^pmrB_") ~ "Colistin",  
      
      # === 13. Glycopeptides ===
      str_detect(AMR.genotypes, "^van") ~ "Glycopeptide",
      
      # === 14. Rifamycins ===
      str_detect(AMR.genotypes, "^arr") ~ "Rifamycin",
      
      # === 15. Bleomycin resistance ===
      str_detect(AMR.genotypes, "^ble") ~ "Bleomycin",  
      
      # === 16. Streptothricin resistance ===
      str_detect(AMR.genotypes, "^sat") ~ "Streptothricin",  

      # === 17. Streptogramin/Virginiamycin resistance ===
      str_detect(AMR.genotypes, "^vat") ~ "Streptogramin", 
      
      # === 18. Multidrug Efflux Pumps (RND family + tmex/toprJ) ===
      str_detect(AMR.genotypes, "^mds|^acr|^mex|^ade|^cme|^amrB|^tmex|^toprJ") ~ "Efflux Pump",
      
      # === 19. MDR Transcriptional Regulators (upregulate efflux) ===
      str_detect(AMR.genotypes, "^ramR_|^soxR_|^marR|^rob") ~ "MDR Regulator",

      # === 20. Gentamicin ===
      str_detect(AMR.genotypes, "grd") ~ "Gentamicin",
           
      # === 21. Default ===
      TRUE ~ "Other"
    )
  ) 

```

```{r}
# Splitting AMR genes and their status

amr_long <- amr_long %>%
   mutate(
    AMR_gene = str_extract(AMR.genotypes, "^[^=]+"),  # e.g., "aacA56=COMPLETE" → "aacA56"
    AMR_status = str_extract(AMR.genotypes, "(?<==).*") #it keeps the information after "="
   )   

#Removing AMR genotypes after split
amr_long$AMR.genotypes	=	NULL 

#Saving the new data
write_csv(amr_long, here("data/processed-data/", "amr_long.csv"))
```

# AMR summaries
```{r}
# Gene prevalence overall
gene_prev <- amr_long %>%
  group_by(AMR_gene, AMR_status) %>%
  summarise(n = n(), .groups = "drop") 

write_csv(gene_prev, here("data/processed-data/", "gene_prev.csv"))

gene_prev

# Class prevalence overall
class_prev <- amr_long %>%  
  filter(AMR_status == "COMPLETE") %>%  
  group_by(AMR_class) %>%  
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%  
  arrange(desc(n)) %>%  
  mutate(
    Percentage = round(n / totaln_isolates * 100, 1) 
  )

class_prev

# Plotting classes frequencies
ggplot(class_prev, aes(x = fct_reorder(AMR_class, n), y = n)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw()
```

##### The vast majority of AMR predicted genes are efflux pumps. I have decided to delete them because it would influence the overall AMR analysis. ¨Several studies show that overexpression of efflux pumps rarely provides clinical resistance but contributes to a low-level resistance.¨ https://doi.org/10.3390/antibiotics9120855 Therefore, while technically AMR, it is so widespread that it should not be considered as a factor for higher risk of developing AMR if already over 90% of isolates show to have it.

```{r}
# Removing Efflux pumps
amr_long1 <- amr_long[!(amr_long$AMR_class %in% "Efflux Pump"), ]

# Keeping only completed profiles
amr_long1 <- amr_long1[(amr_long1$AMR_status %in% "COMPLETE"), ]

# Class prevalence after removing Efflux pumps and keeping only completed profiles
class_prev1 <- amr_long1 %>%
  group_by(AMR_class) %>%  
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%  
  arrange(desc(n)) %>%  
  mutate(
    Percentage = round(n / totaln_isolates * 100, 1)          
  )

class_prev1

write_csv(class_prev1, here("data/processed-data/", "class_prev1csv"))

# Plotting classes frequencies after removing Efflux pumps and keeping only completed profiles
ggplot(class_prev1, aes(x = fct_reorder(AMR_class, n), y = n)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw()

#Saving file
ggsave(filename = "amr_graph1.png", path = here("results/figures/"))

```

#### This indicates that most prevalent drug resistant classes are in Salmonella are: Tetracycline, Aminoglycoside, Sulfonamide, Beta-lactam, Phenicol, Quinolone, Trimethoprim, Fosfomycin. 

```{r}
# Create the count of AMR classes
amr_count <- amr_long1 %>%
  filter(AMR_gene != "None") %>%
  distinct(BioSample, AMR_class) %>% 
  count(BioSample, name = "n_classes")

# Join and Mutate to original dataset
isolates1 <- isolates1 %>%
  left_join(amr_count, by = "BioSample") %>%
  mutate(n_classes = tidyr::replace_na(n_classes, 0))

amr_nclasses <- isolates1 %>%
  group_by(n_classes) %>%
  summarise(n = n(), .groups = "drop") 

amr_nclasses

#Creating a variable based on n_classes 
isolates1 <- isolates1 %>%
  mutate(
    MDR_category = case_when(
      n_classes == 0 ~ "0",
      n_classes == 1 ~ "1",
      n_classes == 2 ~ "2",
      n_classes >= 3 ~ "3+",
    ),
    MDR_category = factor(MDR_category, levels = c("0", "1", "2", "3+"), ordered = TRUE)
  )

# Plot AMR by number of classes
  ggplot(isolates1, aes(x = MDR_category)) + 
  geom_bar() +
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free")

ggsave(filename = "amr_graph2.png", path = here("results/figures/"))


```

##### While the scales are different, it is interesting to observe that the trends between environmental and human AMR look very familiar. While the AMR between animal and food resemble also look similar to each other. Proportionally, there are more isolates with MDR observed in animal and food sources. However, we have to take in consideration this is only observational data.

```{r}
# Creating a wide version for understanding cooccurrance
amr_wide <- amr_long1 %>%
  filter(AMR_gene != "None") %>%
  # Keep only the columns needed for the pivot + identifiers
  select(BioSample, AMR_class, serotype, Source.type) %>%
  # Remove duplicate BioSample + AMR_class combinations while preserving metadata
  distinct(BioSample, .keep_all = TRUE) %>%
  mutate(present = 1) %>%
  pivot_wider(
    id_cols = c(BioSample, serotype, Source.type),
    names_from = AMR_class,
    values_from = present,
    values_fill = 0
  )

```

```{r}
amr_order <- c(
  "Tetracycline", "Aminoglycoside", "Sulfonamide", "Beta_lactam", "Phenicol", "Quinolone", "Trimethoprim", "Fosfomycin", "Macrolide", "Lincosamide", "Rifamycin", "Bleomycin", "Colistin", "Gentamicin", "Streptothricin", "Glycopeptide"
)

#AMR matrix full dataset
amr_mat <- amr_wide %>%
  select(-BioSample) %>%   
  select(where(is.numeric)) %>%
  as.matrix()

#Correlation matrix
cor_mat <- cor(amr_mat)
cor_mat

cor_df <- as.data.frame(as.table(cor_mat)) %>%
  rename(AMR1 = Var1, AMR2 = Var2, corr = Freq) %>%
  mutate(
    AMR1 = factor(AMR1, levels = amr_order),
    AMR2 = factor(AMR2, levels = amr_order)
  ) %>%
    drop_na()


ggplot(cor_df, aes(AMR1, AMR2, fill = corr)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "AMR Correlation Heatmap")
```

##### From this information we can tell the cooccurrance of AMR genes. For the three biggest categories
##### > Tetracycline has its strongest relationship with Aminoglycosides and Sulfonamides
##### > Aminoglycosides has its strongest relationship with Tetracycline and Beta_lactams
##### > Sulfonamides has its strongest relationship with Aminoglycosides, Tetracycline, and Beta_lactams
##### Quinolone, Fosfomycin, MDR Regulator, and Streptogramin seem to have a negative correlation with most of the other classes.
##### > In particular, Quinolone and Fosfomycin seem to most negative correlation between each other

##### After learning about the MDR and the coocurrance af AMR in all isolates, I am interested in two more things: Is cooccurance the same across sources? and what are the serovars leading each one of the main AMR classes?

```{r}
# ======= AMR matrix per source: HUMAN =========== #
amr_mat_human <- amr_wide %>%
  filter(Source.type == "human") %>%
  select(-BioSample) %>%   
  select(where(is.numeric)) %>%
  as.matrix()

#Correlation matrix
cor_mat_human <- cor(amr_mat_human) 
cor_mat_human

cor_df_human <- as.data.frame(as.table(cor_mat_human)) %>%
  rename(AMR1 = Var1, AMR2 = Var2, corr = Freq) %>%
  mutate(
    AMR1 = factor(AMR1, levels = amr_order),
    AMR2 = factor(AMR2, levels = amr_order)
  ) %>%
    mutate(corr = replace_na(corr, 0))
    #drop_na()

ggplot(cor_df_human, aes(AMR1, AMR2, fill = corr)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "AMR Correlation Heatmap: HUMAN")

# ======= AMR matrix per source: ENVIRONMENTAL =========== #
amr_mat_env <- amr_wide %>%
  filter(Source.type == "environmental") %>%
  select(-BioSample) %>%   
  select(where(is.numeric)) %>%
  as.matrix()

#Correlation matrix
cor_mat_env <- cor(amr_mat_env) 
cor_mat_env

cor_df_env <- as.data.frame(as.table(cor_mat_env)) %>%
  rename(AMR1 = Var1, AMR2 = Var2, corr = Freq)   %>%
  mutate(
    AMR1 = factor(AMR1, levels = amr_order),
    AMR2 = factor(AMR2, levels = amr_order)
  ) %>%
    mutate(corr = replace_na(corr, 0))
    #drop_na() %>%
    #filter(!(AMR1 %in% c("Glycopeptide", "Streptothricin")))

ggplot(cor_df_env, aes(AMR1, AMR2, fill = corr)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "AMR Correlation Heatmap: ENVIRONMENTAL")

# ======= AMR matrix per source: ANIMAL =========== #
amr_mat_animal <- amr_wide %>%
  filter(Source.type == "animal") %>%
  select(-BioSample) %>%   
  select(where(is.numeric)) %>%
  as.matrix()

#Correlation matrix
cor_mat_animal <- cor(amr_mat_animal)
cor_mat_animal

cor_df_animal <- as.data.frame(as.table(cor_mat_animal)) %>%
  rename(AMR1 = Var1, AMR2 = Var2, corr = Freq) %>%
  mutate(
    AMR1 = factor(AMR1, levels = amr_order),
    AMR2 = factor(AMR2, levels = amr_order)
  ) %>%
    mutate(corr = replace_na(corr, 0))
    #drop_na() %>%
    #filter(!(AMR1 %in% c("Glycopeptide", "Streptothricin")))

ggplot(cor_df_animal, aes(AMR1, AMR2, fill = corr)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "AMR Correlation Heatmap: ANIMAL")

# ======= AMR matrix per source: FOOD =========== #
amr_mat_food <- amr_wide %>%
  filter(Source.type == "food") %>%
  select(-BioSample) %>%   
  select(where(is.numeric)) %>%
  as.matrix()

#Correlation matrix
cor_mat_food <- cor(amr_mat_food, use = "pairwise.complete.obs") 
cor_mat_food

cor_df_food <- as.data.frame(as.table(cor_mat_food)) %>%
  rename(AMR1 = Var1, AMR2 = Var2, corr = Freq) %>%
  mutate(
    AMR1 = factor(AMR1, levels = amr_order),
    AMR2 = factor(AMR2, levels = amr_order)
  ) %>%
    mutate(corr = replace_na(corr, 0))
    #drop_na()

ggplot(cor_df_food, aes(AMR1, AMR2, fill = corr)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "AMR Correlation Heatmap: FOOD")
```

###### From this information we can observe that trends are different among sources. Animal and environmental matrixes resemble closely the correlations observed in human isolates. Also, only human isolates presented Glycopeptide and Streptothricin as predicted AMR classes. 

##### Additional observations:
###### > The pairs: Gentamicin-Colistin and Rifamycin-Colistin are both infrequent but they seem to cooccur in animal and environmental samples. 
###### > Tetracycline, Aminoglycoside, Sulfonamide, Beta_lactams, and Phenicol are high frequency and cooccurance with each other in human, animal, and environmental samples
###### > Food isolates have the lowest correlation values among AMR classes
###### > Fosfomycin has a negative correlation to the majority of AMR classes across all sources

```{r}
#Plotting top serovars
amr_long1 %>% 
  group_by(serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%  
  arrange(desc(n)) %>% 
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = fct_reorder(serotype, n, .desc = FALSE, .fun = max), 
             y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  theme(legend.position = "none")

#Plotting top serovars by each source
amr_long1 %>% 
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none")

```

```{r}
amr_long1 %>%
  filter(serotype == "I 4,[5],12:i:-") %>%
  group_by(Source.type, AMR_class) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  ggplot(aes(x = fct_reorder(AMR_class, n), y = n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free_x", nrow = 1) +
  labs(title = "I 4,[5],12:i:-")

amr_long1 %>%
  filter(serotype == "Newport") %>%
  group_by(Source.type, AMR_class) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  ggplot(aes(x = fct_reorder(AMR_class, n), y = n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free_x", nrow = 1) +
  labs(title = "Newport")

amr_long1 %>%
  filter(serotype == "Typhimurium") %>%
  group_by(Source.type, AMR_class) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  ggplot(aes(x = fct_reorder(AMR_class, n), y = n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free_x", nrow = 1) +
  labs(title = "Typhimurium")

amr_long1 %>%
  filter(serotype == "Infantis") %>%
  group_by(Source.type, AMR_class) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  ggplot(aes(x = fct_reorder(AMR_class, n), y = n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free_x", nrow = 1) +
  labs(title = "Infantis")

amr_long1 %>%
  filter(serotype == "Dublin") %>%
  group_by(Source.type, AMR_class) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  ggplot(aes(x = fct_reorder(AMR_class, n), y = n)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free_x", nrow = 1) +
  labs(title = "Dublin")

```

```{r}
amr_long1 %>% 
  filter(AMR_class == "Tetracycline") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Tetracycline")
```

```{r}
#Now serovar trends will be graph per each one of the AMR classes, while not all of them will be included in the manuscruipt. Maybe as supplemental material.

amr_long1 %>% 
  filter(AMR_class == "Tetracycline") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Tetracycline")

amr_long1 %>% 
  filter(AMR_class == "Aminoglycoside") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Aminoglycoside")

amr_long1 %>% 
  filter(AMR_class == "Sulfonamide") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Sulfonamide")

amr_long1 %>% 
  filter(AMR_class == "Beta_lactam") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Beta_lactam")

amr_long1 %>% 
  filter(AMR_class == "Phenicol") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Phenicol")

amr_long1 %>% 
  filter(AMR_class == "Quinolone") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Quinolone")

amr_long1 %>% 
  filter(AMR_class == "Trimethoprim") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Trimethoprim")

amr_long1 %>% 
  filter(AMR_class == "Fosfomycin") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Fosfomycin")

amr_long1 %>% 
  filter(AMR_class == "Macrolide") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Macrolide")

amr_long1 %>% 
  filter(AMR_class == "Lincosamide") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Lincosamide")

amr_long1 %>% 
  filter(AMR_class == "Rifamycin") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Rifamycin")

amr_long1 %>% 
  filter(AMR_class == "Bleomycin") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Bleomycin")

amr_long1 %>% 
  filter(AMR_class == "Colistin") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Colistin")

amr_long1 %>% 
  filter(AMR_class == "Gentamicin") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Gentamicin")

amr_long1 %>% 
  filter(AMR_class == "Streptothricin") %>%
  group_by(Source.type, serotype) %>%
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%
  group_by(Source.type) %>%
  slice_max(n, n = 5, with_ties = FALSE) %>% 
  ungroup() %>%
  ggplot(aes(x = reorder_within(serotype, n, Source.type), y = n, fill = serotype)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free") +
  scale_x_reordered() +
  theme(legend.position = "none") +
  labs(title = "Streptothricin")

```

