---
title: "Descriptive analysis script by Marco Reina"
output: html_document
---
#### Code originally written by MR and improve with ChatGPT & Qween

```{r}
# Load libraries
library(tidyverse)
library(readr)
library(here)
library(dplyr)
library(stringr)
library(knitr)
library(forcats) 

# Read data
isolates1 <- read_csv(here("data/processed-data/", "isolates1.csv"))

#Split computed serovars
isolates1 <- isolates1 %>%
  mutate(
    antigen_formula = str_extract(Computed.types, "(?<=antigen_formula=).*?(?=,serotype=)"),
    serotype = str_extract(Computed.types, "(?<=serotype=).*$")
  )

# Removing columns that are no longer useful after the original cleaning
isolates1$Computed.types	=	NULL
isolates1$Location	=	NULL
isolates1$country_name	=	NULL 
```

##### The goal of this analysis is to characterize AMR incidence in _Salmonella_
##### As the data comes from an observational surveillance program, only descriptive analysis will be performed

```{r}
#Split AMR genes
amr_long <- isolates1 %>%
  # Keep isolates even if no AMR genes
  mutate(AMR.genotypes = if_else(
    is.na(AMR.genotypes) | AMR.genotypes == "",
    "None=NONE",
    AMR.genotypes
  )) %>%
  # Split genes into rows
  separate_rows(AMR.genotypes, sep = ",") 
```

```{r}
# Assign antimicrobial class

amr_long <- amr_long %>%
   mutate(
      AMR_class = case_when(
      # === 1. No AMR ===
      str_detect(AMR.genotypes, "None") ~ "No AMR",
      
      # === 2. Beta-lactams ===
      str_detect(AMR.genotypes, "^bla") ~ "Beta_lactam",
      
      # === 3. Tetracyclines ===
      str_detect(AMR.genotypes, "tet") ~ "Tetracycline",
      
      # === 4. Sulfonamides ===
      str_detect(AMR.genotypes, "^sul") ~ "Sulfonamide",
      
      # === 5. Trimethoprim ===
      str_detect(AMR.genotypes, "^dfr") ~ "Trimethoprim",
      
      # === 6. Quinolones: target mutations & protection proteins ===
      str_detect(AMR.genotypes, "gyrA|gyrB|parC_|parE_|^qnr|^oqx|^qepA") ~ "Quinolone",
      
      # === 7. Aminoglycosides: modifying enzymes (broadened) ===
      str_detect(AMR.genotypes, "aad|aph\\(|aac\\(|armA|^rmt|^ant|^aphA|^aacA") ~ "Aminoglycoside",
      
      # === 8. Phenicols (chloramphenicol family) ===
      str_detect(AMR.genotypes, "floR|^cat|^cml|^gar") ~ "Phenicol",  
      
      # === 9. Macrolides: including esterases (estT/estX) ===
      str_detect(AMR.genotypes, "erm|^mef|^mph|^msr|^ere|^est") ~ "Macrolide",  

      # === 10. Lincosamides ===
      str_detect(AMR.genotypes, "^lnu") ~ "Lincosamide",
      
      # === 11. Fosfomycin ===
      str_detect(AMR.genotypes, "^fos") ~ "Fosfomycin",
      
      # === 12. Colistin/Polymyxin: mcr enzymes + pmr regulators ===
      str_detect(AMR.genotypes, "^mcr|^pmrA_|^pmrB_") ~ "Colistin",  
      
      # === 13. Glycopeptides ===
      str_detect(AMR.genotypes, "^van") ~ "Glycopeptide",
      
      # === 14. Rifamycins ===
      str_detect(AMR.genotypes, "^arr") ~ "Rifamycin",
      
      # === 15. Bleomycin resistance ===
      str_detect(AMR.genotypes, "^ble") ~ "Bleomycin",  
      
      # === 16. Streptothricin resistance ===
      str_detect(AMR.genotypes, "^sat") ~ "Streptothricin",  

      # === 17. Streptogramin/Virginiamycin resistance ===
      str_detect(AMR.genotypes, "^vat") ~ "Streptogramin", 
      
      # === 18. Multidrug Efflux Pumps (RND family + tmex/toprJ) ===
      str_detect(AMR.genotypes, "^mds|^acr|^mex|^ade|^cme|^amrB|^tmex|^toprJ") ~ "Efflux Pump",
      
      # === 19. MDR Transcriptional Regulators (upregulate efflux) ===
      str_detect(AMR.genotypes, "^ramR_|^soxR_|^marR|^rob") ~ "MDR Regulator",

      # === 20. Gentamicin ===
      str_detect(AMR.genotypes, "grd") ~ "Gentamicin",
           
      # === 21. Default ===
      TRUE ~ "Other"
    )
  ) 

```

```{r}
# Splitting AMR genes and their status

amr_long <- amr_long %>%
   mutate(
    AMR_gene = str_extract(AMR.genotypes, "^[^=]+"),  # e.g., "aacA56=COMPLETE" → "aacA56"
    AMR_status = str_extract(AMR.genotypes, "(?<==).*") #it keeps the information after "="
   )   

#Removing AMR genotypes after split
amr_long$AMR.genotypes	=	NULL 

#Saving the new data
write_csv(amr_long, here("data/processed-data/", "amr_long.csv"))
```

# AMR summaries
```{r}
# Gene prevalence
gene_prev <- amr_long %>%
  group_by(AMR_gene, AMR_status) %>%
  summarise(n = n(), .groups = "drop") 

write_csv(gene_prev, here("data/processed-data/", "gene_prev.csv"))

gene_prev

# Class prevalence
class_prev <- amr_long %>%  
  filter(AMR_status == "COMPLETE") %>%  
  group_by(AMR_class) %>%  
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%  
  arrange(desc(n)) %>%  
  mutate(
    pct_total = round(n / sum(n) * 100, 1),           
    cum_pct = round(cumsum(pct_total), 1)         
  )

write_csv(class_prev, here("data/processed-data/", "class_prev.csv"))

class_prev

# Plotting classes frequencies
ggplot(class_prev, aes(x = fct_reorder(AMR_class, n), y = n)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw()
```

##### The vast majority of AMR predicted genes are efflux pumps. I have decided to delete them because it would influence the overall AMR analysis. ¨Several studies show that overexpression of efflux pumps rarely provides clinical resistance but contributes to a low-level resistance.¨ https://doi.org/10.3390/antibiotics9120855 Therefore, while technically AMR, it is so widespread that it should not be considered as a factor for higher risk of developing AMR if already over 90% of isolates show to have it.

```{r}
# Removing Efflux pumps
amr_long1 <- amr_long[!(amr_long$AMR_class %in% "Efflux Pump"), ]

# Class prevalence after removing Efflux pumps
class_prev1 <- amr_long1 %>%
  filter(AMR_status == "COMPLETE") %>%  
  group_by(AMR_class) %>%  
  summarise(n = n_distinct(BioSample), .groups = "drop") %>%  
  arrange(desc(n)) %>%  
  mutate(
    pct_total = round(n / sum(n) * 100, 1),           
    cum_pct = round(cumsum(pct_total), 1)         
  )

class_prev1

# Plotting classes frequencies after removing Efflux pumps
ggplot(class_prev1, aes(x = fct_reorder(AMR_class, n), y = n)) + 
  geom_col() +  
  coord_flip() +  
  theme_bw()

#Saving file
ggsave(filename = "amr_graph1.png", path = here("results/figures/"))
```


#### This indicates that ~95% of drug resistant classes are: Tetracycline, Aminoglycoside, Sulfonamide, Beta-lactam, Phenicol, Quinolone, Trimethoprim, Fosfomycin. 

```{r}
# Create the count of AMR classes
amr_count <- amr_long1 %>%
  filter(AMR_gene != "None") %>%
  distinct(BioSample, AMR_class) %>% 
  count(BioSample, name = "n_classes")

# Join and Mutate to original dataset
isolates1 <- isolates1 %>%
  left_join(amr_count, by = "BioSample") %>%
  mutate(n_classes = tidyr::replace_na(n_classes, 0))

amr_nclasses <- isolates1 %>%
  group_by(n_classes) %>%
  summarise(n = n(), .groups = "drop") 

amr_nclasses

#Creating a variable based on n_classes 
isolates1 <- isolates1 %>%
  mutate(
    MDR_category = case_when(
      n_classes == 0 ~ "0",
      n_classes == 1 ~ "1",
      n_classes == 2 ~ "2",
      n_classes >= 3 ~ "3+",
    ),
    MDR_category = factor(MDR_category, levels = c("0", "1", "2", "3+"), ordered = TRUE)
  )

# Plot AMR by number of classes
  ggplot(isolates1, aes(x = MDR_category)) + 
  geom_bar() +
  theme_bw() +
  facet_wrap(~ Source.type, scales = "free")

ggsave(filename = "amr_graph2.png", path = here("results/figures/"))

```

##### While the scales are different, it is interesting to observe that the trends between environmental and human AMR look very familiar. While the AMR between animal and food resemble also look similar to each other.

```{r}
# Creating a wide version for understanding cooccurrance
amr_wide <- amr_long1 %>%
  filter(AMR_gene != "None") %>%
  distinct(BioSample, AMR_class) %>%  
  mutate(present = 1) %>%
  pivot_wider(
    id_cols = BioSample,
    names_from = AMR_class,
    values_from = present,
    values_fill = 0
  )

#Removing classes that represent less than 5%
amr_wide$Macrolide = NULL
amr_wide$Lincosamide = NULL
amr_wide$Rifamycin = NULL
amr_wide$Bleomycin = NULL
amr_wide$Colistin = NULL
amr_wide$Gentamicin = NULL
amr_wide$Streptothricin = NULL
amr_wide$Glycopeptide = NULL

isolates2 <- left_join(isolates1, amr_wide, by = "BioSample") %>%
   mutate(across(where(is.numeric), ~replace_na(.x, 0)))

isolates2$AMR.genotypes = NULL
```

```{r}
#AMR matrix
amr_mat <- amr_wide %>%
  select(-BioSample) %>%   # keep only features
  select(where(is.numeric)) %>%
  as.matrix()

#Correlation matrix
cor_mat <- cor(amr_mat)
cor_mat

cor_df <- as.data.frame(as.table(cor_mat)) %>%
  rename(AMR1 = Var1, AMR2 = Var2, corr = Freq)

ggplot(cor_df, aes(AMR1, AMR2, fill = corr)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "AMR Correlation Heatmap")
```

##### From this information we can tell the cooccurrance of AMR genes. For the three biggest categories
##### > Tetracycline has its strongest relationship with Aminoglycosides and Sulfonamides
##### > Aminoglycosides has its strongest relationship with Tetracycline and Beta_lactams
##### > Sulfonamides has its strongest relationship with Aminoglycosides, Tetracycline, and Beta_lactams
##### Quinolone, Fosfomycin, MDR Regulator, and Streptogramin seem to have a negative correlation with most of the other classes.
##### > In particular, Quinolone and Fosfomycin seem to most negative correlation between each other

```{r}
# Option 2, Jaccard matrix. Not sure if going to be used, most likely Pearson Correlation will be used (above)
cooc_mat <- t(amr_mat) %*% amr_mat

cooc_df <- as.data.frame(as.table(cooc_mat)) %>%
  rename(AMR1 = Var1, AMR2 = Var2, count = Freq) %>%
  filter(AMR1 != AMR2)

row_sums <- diag(cooc_mat)

jaccard <- cooc_mat / (outer(row_sums, row_sums, "+") - cooc_mat)

jaccard
```



